Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-06-03 paper1_jiturra.Rmd
ediciones v2

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | Escalating economic inequality has increased attention to the link between class-based social ties and redistributive preferences in contemporary societies. However, social classes are not equally affected by inequality regarding their chances of establishing cross-class social relations. While those in the upper and middle classes have greater access to economic and diverse social ties, the working class has been disproportionately affected by social exclusion. Consequently, this leads to a growing divide between their lives and the lives of others, consolidating segregated lifeworlds. Previous research has shown that the link between class relations and redistributive preferences can be enriched from a network perspective. Meeting people from diverse class positions can help individuals become more aware of the lifestyles and worldviews of others. On the one hand, when upper-class networks become more homogeneous, it can lead to lower empathy toward those in need and less support for redistributive policies. On the other hand, when there is segregation in lower-class social networks, it can lead to greater marginalization and a higher demand for welfare support. This study explores how class-based network segregation influences redistributive preferences from a cross-national perspective, employing data from the International Social Survey 2017, which comprised 32,529 individuals in 31 societies. The regression analyses indicate that network homogeneity has a conditional influence on redistributive preferences based on social class. Homogeneous social networks in the working and intermediate classes increase redistributive preferences, while greater homogeneity in the upper classes reduces support for redistribution. In addition, in contexts of greater income inequality, the conditional influence of network segregation weakens, especially for the upper classes. Implications for the study of class relations and policy preferences are discussed.
  | 
  | **Keywords**: social networks, social class, redistributive preferences, income inequality
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa6.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

@@

```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")  #si falta pacman, instalar
if (!require("tinytex")) install.packages("tinytex")#si falta tinytex, instalar
pacman::p_load(knitr, kableExtra, dplyr, ggplot2,sjmisc,texreg) # librerias
knitr::opts_chunk$set(warning = FALSE,  # mensaje de warning
                      message = FALSE,  # mensajes/avisos de librerias  
                      cache = F,    # cache de los chunks,usar analisis pesados
                      out.width = '100%',
                      fig.pos= "H",     # posicion figuras H = HERE
                      fig.align = "center",
                      echo = FALSE      # incluir chunk en output
                      )
options(scipen=999) # notacion cientifica
rm(list=ls())       # limpiar workspace
options(knitr.kable.NA = '') # NA en kable = ''
```
\pagestyle{fancy}
\fancyhead[R]{\thepage}
\fancyhead[L]{Research Article Draft}

@@

The study of redistributive preferences and social class has been mainly focused on the individual and household situation. Despite this, less attention has been paid to the role of social networks as socialization agents of attitudes toward redistribution. It has been stated that friendship and partnership ties tend to be homogeneous in their characteristics, expressing the tendency to form and consolidate homophilous social networks [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Thus, the scope of class analysis in attitude formation urges to be extended to stress the role of social ties and segregation as they involve social processes, such as class identity formation, internalization of shared norms, conflict, and cooperation between classes [@Svallfors2006]. In this respect, recent research has argued that the degree of segregation in the composition of social ties can influence opinions about economic inequality as they function as a window for learning about the lifestyles, economic conditions, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. However, there is still a knowledge gap to disentangle the network mechanisms involved in political attitude formation about the role of government in enacting redistributive policies to strive against economic inequality [@mccall_americans_2009].

The class composition of network ties can influence opinions on redistributive policies, but also the degree of segregation from other social classes in more homogeneous social networks can strengthen the class divide in redistributive preferences  [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention should be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003].

Besides the individual dynamics, little is known about how contextual factors affect the relationship between networks and redistributive preferences. Previous studies have shown that wider economic gaps lead to segregated forms of social participation and status composition of social networks [@pichler_patterns_2007;@otero_space_2022]. Indeed, economic inequality reinforces the stratified access to social activities and the social distance between social classes, resulting in an increasing marginalization of the lower classes and consolidating the privileged positions of the upper classes that held better opportunities for diversifying their social activities and interpersonal ties [@lancee_income_2012;@letki_getting_2015;@otero_differences_2023]. Studies on the class-attitude link posit that economic inequality is crucial for understanding how class relations are reflected in redistributive demands as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. The main finding drawn from these studies is that the upper classes tend to hold more egalitarian attitudes in contexts of high inequality, while the attitudes of the working class remain stable regardless of inequality levels. In this discussion, the moral economy explanation argues that the upper classes are more concerned about the integrity of resource allocation procedures as a matter of distributive justice [@sachweh_why_2019]. Conversely, political economy research suggests that the upper classes are more likely to support redistribution policies due to the negative consequences of inequality [@dimick_altruistic_2017].

In this context, to our knowledge, no research has comparatively stressed the role of economic inequality in the relationship between network segregation and redistributive preferences. Therefore, the present study aims to answer two research questions. First, at the micro-level, what is the role of class-based network segregation on redistributive preferences? Secondly, at the macro level, to what extent does economic inequality condition the relationship between class-based network segregation and redistributive preferences? To address these questions, we have analyzed data from the International Social Survey Program (ISSP) 2017, which includes a sample of 32,529 individuals across 31 societies. In this regard, the ISSP presents a nonprecedence opportunity as it provides comprehensive information on interpersonal networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of redistributive preferences in industrialized societies has consistently demonstrated the importance of social class as a key explanatory factor  [@lindh_class_2020, p. 421]. Specifically, class positions have been shown to significantly influence individual opinions on the government's role in various aspects of the economic sphere, such as reducing income disparities or enhancing opportunities for those in economic distress [@mccall_americans_2009].

However, class-based explanations of redistributive preferences have mainly been, but not exclusively, focused on the individual or household situation. For instance, it has been claimed that redistributive preferences can be explained through the labor market situation, which comprises access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009].
Moreover, as material interests might prevail in scarcity, value-driven motivations are a relevant explanatory factor of redistributive preferences [@Kulin2013;@feldman_humanitarian_2001], becoming salient under greater certainty and weaker under material precariousness [@Maldonadoetal2019].Another group of studies has suggested that given the substantial time workers spend performing their jobs, social relations in the workspace can also imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. It is argued that continuous interaction with diverse people that characterize interpersonal services can reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, the vertical monitoring in managerial occupations and the relevance of autonomy in the case of self-employed can contribute to self-interested and conservative political views [@langsaether_more_2020;@oesch_electoral_2018].

While our understanding of how social class influences public opinion on market-based resource distribution [@andersen_preferences_2018;@lindh_public_2015] and government income redistribution [@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010;@Svallfors2006] is substantial, there remains a significant gap in our knowledge about the impact of class-based social relations from a network perspective on redistributive preferences.

@@

It is possible to argue that class relations represent not only resource-based distinctions but also patterns of sociability and exchange observed in the differentiation of social ties. Class relations can also be understood as the degree of cross-class relationship as a structural characteristic of a society, representing a network of social ties between different class positions [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001, p.416]. Here, friendship and family tend to be homogeneous in terms of their demographic characteristics that simultaneously intersect with socioeconomic status and class [@bargsted_pautas_2020; @plaza_juntos_2022;@lazarsfeld_friendship_1954]. Acknowledging that socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004;@homans_human_1951], we argue that attitudinal similarity is the result of segregated social relations, which in turn are strongly linked to structural processes of differentiation that ultimately demarcate the opportunities in the formation of diverse social ties [@feld_focused_1981].

In the discussion on the class composition of social networks, it is possible to find two approaches. On the one hand, one approach has emphasized the role of diversity as the degree of connectedness to dissimilar occupations as social positions that vertically represent access to resources embedded in social networks [@lin_social_2007]. However, diversity is defined as the rate of dissimilar ties or the total ties to certain groups that do not necessarily count with a reference position to describe the network composition. On the other hand, a second approach has underscored the role of network segregation, which has been described as lacking cross-class network ties and is conceptually more proximate to the homophily principle because it is anchored in individual class positions [@otero_open_2021].

Empirically, studies on network diversity have shown that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the more homogeneous participation of the working class [@pichler_social_2009]. Similar patterns have been found in terms of the composition of social ties, where the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes [@cepic_how_2020;@carrascosa_class_2023]. Also, this segregation pattern holds during the life course, where the upper classes hold increasingly diverse social contacts in contrast to the stable networks of the working classes [@volker_social_2020]. In contrast, studies that have focused on segregation suggest that the property dimension is much less permeable than authority-based boundaries, given that class interests increase the social distance between proprietors and workers, while the intermediate class position of supervisors and contact frequency with workers make friendship tie formation more likely [@wright_relative_1992] Similar studies have shown that the intermediate classes' higher permeability contrasts with the working class's more homogeneous networks, suggesting that their limited resources and lower capacity to be socially engaged ultimately result in a lack of social resources that lead to social segregation. In contrast, the upper class is less permeable and homogeneous because it tends to self-select as a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

In addition to individual-based explanations, it is argued that people form opinions in the economic domain based on their social relations. Despite limited research on the link between social networks and redistributive preferences, I identify two broad theoretical approaches that have discussed the role of the social environment in attitude formation.

On the one hand, one approach accentuates the role of beliefs about economic inequality rooted in social comparison processes as an explanation through which people form opinions about redistributive policies [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. Here, the initial argument posits that people form their beliefs through the individual, family, friends, and coworkers’ experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of reference groups [@Evans1992, p. 467]. From this perspective, how people infer the social world is linked to the degree of segregation in their immediate social environment, which influences the intensity and character of the information that ultimately shapes their perceptions of inequality [@mijs_is_2021]. Thus, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more segregated networks [@summers_deliberating_2022]. Despite this, we argue that this research field has been more focused on the cognitive dimension of preference formation through inequality perceptions, either relying on surveys [@cansunar_who_2021;@garcia-castro_perceived_2022] or experimental manipulations [@Cruces2011;@becker_significant_2021] rather than empirically addressing the role of class segregation in social networks and its claimed influence on attitudes in the economic domain.

In contrast, it has been suggested that the lens of a network approach provides a better picture of class relations that nurture social norms and group identity [@kalmijn_social_2007, p. 550]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to individual or household material well-being, cultural perspectives, and political preferences that structure broader social experiences [@morris_attenuation_1996, p. 48]. Furthermore, social integration can be affected in societies with lower contact opportunities between different social classes, where social distance among them widened by extreme inequality can create an “empathy gulf” that comprises barriers to imagining others’ lifestyles [@sachweh_moral_2012]. In this matter, spatially segregated interactions may nurture doubts about worthiness when the lower classes contrast themselves with the lifestyles of the upper classes, which might undermine feelings of social inclusion and cohesion [@sachweh_moral_2012]. Consequently, segregation drives the lives of others to become more distant and might have consequences for empathy and solidarity toward others, potentially leading to the perception of fellow citizens as strangers [@otero_lives_2022, p. 758]. Thus, we suggest that attitudes towards redistribution may be influenced by the class situation of both the individual and network ties, where social influence processes can either allineate or divide attitudes according to the characteristics of groups and the level of segregation that allows factual contact opportunities.

In recent years, we have seen an increase in this network turn of class analysis, contributing to unraveling the relationship between the composition of social ties and attitudes toward inequality. In this regard, it has been shown that the degree of contact between different classes has shown that individuals of the lower class interacting with those of the upper classes may consider economic inequality as justifiable, whereas interactions with those of the lower class may prompt individuals from different social classes to question the fairness of income distribution [@vargassalfate_contact_2023]. Further evidence suggests that class-based contact diversity shortens the social distance between social classes, providing broader images of the living conditions of others, which can trigger greater concerns about inequality [@otero_power_2023] and undermine support for the market distribution of social services [@beck_explorando_2019].

In addition, studies that directly addressed the relationship between networks and social classes have been consistent with previous findings. For instance, evidence from Sweden suggests that people tend to assimilate their opinions according to their surrounding friends and acquaintances, which is reinforced by class homophily, where higher contacts in the managerial class negatively influence support for redistribution, while ties to the working class increase it regardless of the individual class position suggesting that “individuals take an impression from others and modify their attitudes accordingly” [@lindh_missing_2021, p. 698]. Similarly, additional socialization sources are cross-class ties through family members. In this sense, the family of origin is crucial for political socialization, where class interests and norms are nurtured in childhood and early adulthood. A study in the United States showed that those tied to the upper class through parental relations support redistribution and progressive taxation less than those with working-class family backgrounds [@lee_consider_2023]. In addition, @paskov_crossclass_2022 argues that the class position of family and partners shapes preferences, while households share risk according to the class position of their members, where class homogeneity in ties divides preferences between lower and upper classes. In contrast, this gradient is blurred as heterogeneity increases according to the class position of the partner and family of origin.

In summary,  the tendency to form homophilous social ties regarding social class is expected to reinforce attitudes and consolidate opinion similarity. Empirically, a weak direct association between homogeneity and redistributive preferences is expected because class segregation does not distinguish between classes as it provides only the _overall_ degree of segregation. In contrast, the focus will be on how class-based network homogeneity is conditional to social class. In simpler terms, this is possible by examining the interaction between network homogeneity and social class. Precisely, this will indicate if being segregated into lower (upper) classes leads to stronger (weaker) redistributive preferences ($H_1$).

## Economic Inequality as context for Class relations and Redistributive Preferences

>>>>

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
#   number: 1
# - name: Justin d'Ottawa
#   affiliation: University of Ottawa
#   number: 2
# - name: Pedro Torres
#   number: 1,2 
# Nota: Autores que comparten filiacion, poner el mismo number y declarar filiación una sola vez.  
abstract: | 
  | Escalating economic inequality has increased attention to the link between class relations and redistributive preferences in contemporary societies. Hoewever, economic inequality unevenly affects class relations and their influence on support for redistribution across societies. While those in the upper and middle classes reinforce their privileged access to resources and diverse networks, the working class has been disproportionately affected by social exclusion as inequality rises. Consequently, inequality consolidates segregated lifeworlds as it increases the divide between the lives and experiences of the different social classes. In this context, class relations represent the market situation of individuals and households as well as a network of social relations within and between classes. Meeting people from diverse class positions can help individuals become more aware of other people's lifestyles and worldviews. Homogeneous upper-class networks may lead to reduced empathy and solidarity towards those in need, resulting in decreased support for redistribution. Conversely, segregated lower-class social networks can lead to increased marginalization and reinforce support for redistribution. This study explores the relation of class-based network segregation and redistributive preferences employing cross-national data of 32,529 individuals in 31 societies. The main findings suggest that social class conditions the association of network homogeneity on redistributive preferences, where working class homogeneity drive stronger redistributive preferences, while greater upper-class homogeneity decreases support for redistribution. In addition, the conditional influence of network segregation weakens in unequal societies, especially for the upper classes. Implications for the study of class relations and political atttiudes are dicussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa6.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  # - \usepackage{times}           # Times New Roman
  - \usepackage{helvet}
  - \renewcommand{\familydefault}{\sfdefault} 
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

@@

```{r setup, include=FALSE}
if (!require("pacman")) install.packages("pacman")  #si falta pacman, instalar
if (!require("tinytex")) install.packages("tinytex")#si falta tinytex, instalar
pacman::p_load(knitr, kableExtra, dplyr, ggplot2,sjmisc,texreg) # librerias
knitr::opts_chunk$set(warning = FALSE,  # mensaje de warning
                      message = FALSE,  # mensajes/avisos de librerias  
                      cache = T,    # cache de los chunks,usar analisis pesados
                      out.width = '100%',
                      fig.pos= "H",     # posicion figuras H = HERE
                      fig.align = "center",
                      echo = FALSE      # incluir chunk en output
                      )
options(scipen=999) # notacion cientifica
rm(list=ls())       # limpiar workspace
options(knitr.kable.NA = '') # NA en kable = ''
```
\pagestyle{fancy}
\fancyhead[R]{\thepage}
\fancyhead[L]{Research Article Draft}

```{r include=FALSE}
wordcount<- rmdwc::rmdcount(files = "paper1_jiturra.Rmd")
wordcount$words
```

**Word count**:  `r wordcount$words`

@@

During the past decades, the few cross-national studies on redistributive preferences and social class have predominantly focused on the situation of individuals and households [@lindh_class_2020]. Nonetheless, little attention has been paid to the network mechanisms involved in how individuals form their opinions regarding economic inequality and redistributive policies. On the one hand, sociologists have formerly demonstrated that homophily profoundly drives individuals to form similar social ties that consequently nurture social network segregation [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. In contrast, it is also known that opinions about economic inequality can be affected by class segregation as it provides a window for learning about the economic conditions, lifestyles, and worldviews of other social classes [@mijs_inequality_2018;@otero_power_2023]. Despite this, it can be argued that the scrutiny of redistributive preferences can be enhanced by broadening the comprehension of class relations from a network perspective, providing a comprehensive manner to understand processes of shared norm internalization, class identity formation, and collective solidarity [@Svallfors2006;@goldthorpe_sobre_1992]. In this regard, the present study examines the relationship between class-based network segregation and redistributive preferences in cross-national comparison, focusing on the role of economic inequality as a moderator of the claimed influence of class relations on support for redistribution.

While prior research has predominantly focused on examining the impact of social class through an individualistic lens, it is noteworthy that more attention should be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations that make them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. In addition, the normative basis of class relations introduces the relevance of the dimensions of solidarity and reciprocity, which have been argued to provide the moral basis for legitimacy and popular support for welfare schemes [@mau_moral_2003]. In this regard, besides the leverage of individual class position, recent efforts have pointed out that whereas the class positions of network ties do affect redistributive preferences, the degree of segregation in a homogeneous class environment can drive attitudes to become even stronger [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022].

While network segregation intensifies the already established class divide in redistributive preferences, little is known about how contextual factors affect the association of class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link posit that economic inequality is crucial for understanding how class relations are mirrored in redistributive demands as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding on this matter is that the upper classes tend to hold more egalitarian attitudes in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality leads to more segregated forms of social participation and network composition [@pichler_patterns_2007;@otero_space_2022]. Indeed, inequality reinforces the stratified access to social activities and widens the distance between classes, resulting in an increasing marginalization of the lower classes and consolidating the privileged positions of the upper classes that held better opportunities for diversifying their social activities and interpersonal ties [@lancee_income_2012;@letki_getting_2015;@otero_differences_2023]. Nevertheless, current efforts to examine the impact of income inequality have primarily concentrated on either social networks or support for redistribution. A key question that remains unanswered is the extent to which broader economic disparities can influence the relationship between class-based network segregation and support for redistribution within a comprehensive framework.

Thus, as no research has comparatively stressed the role of economic inequality in the relationship between class-based network segregation and redistributive preferences, the present study aims to answer two questions:

(1) How does class-based network segregation affect redistributive preferences? 
(2) To what extent does economic inequality moderate the relationship between class-based network segregation and redistributive preferences?

For the present investigation, a data sample of 32,529 individuals across 31 societies from the International Social Survey Program (ISSP) 2017 is utilized. The ISSP presents a nonprecedence opportunity as it provides comprehensive information on interpersonal networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, the role of class goes beyond the market situation of individuals but drives collective economic interests and moral views about the role of the market and the state [@Svallfors2006]. Here, what has been understood as redistributive preferences entail the public responses to economic inequality regarding the sphere of the role of government actions in aspects such as the reduction of income differences or enhancing opportunities for those in economic despair  [@mccall_americans_2009].

However, class-based explanations of redistributive preferences have mainly been, but not exclusively, focused on the individual or household situation. It has been claimed that redistributive preferences can be explained through the labor market situation, which comprises access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009].
Moreover, as material interests might prevail in scarcity, value-driven motivations are a relevant explanatory factor of redistributive preferences [@Kulin2013;@feldman_humanitarian_2001], becoming salient under greater certainty and weaker under material precariousness [@Maldonadoetal2019]. Other approaches have suggested that given the substantial time workers spend performing their jobs, social relations in the workspace can imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. For instance, continuous and diverse social interactions that characterize interpersonal services can drive empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, the vertical monitoring in managerial occupations and the relevance of autonomy in the case of self-employed fortify self-interested and conservative political views [@langsaether_more_2020;@oesch_electoral_2018].

Empirically, the current understanding of the class divides on redistributive preferences is extensive [@andersen_preferences_2018;@lindh_public_2015; @curtis_how_2015;@langsaether_more_2020;@brooks_why_2010;@Svallfors2006]. Nevertheless, besides the market situation of individuals and their households, including class-based social ties raises the relevance of the _class situation_ as it comprises both economic interests and life chances that motivate communal action beyond the individualistic lens [@weber_class_2011, pp. 57-59]. Therefore, the role of social relations on redistributive preferences from a network perspective needs to be developed beyond mere resource interchanges, but as part of the sociability practices that structure class relations.

@@

It can be argued that class relations not only represent resource-based distinctions but also patterns of sociability observed in the differentiation of social ties. Class relations can also be understood as the degree of cross-class relationship as a structural characteristic of a society, representing a network of social ties between different class positions [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001, p.416]. Hereby, friendship and family ties are prone to be homogeneous in terms of class and demographic characteristics, but distant ties play the role of bridging individuals to other different social groups and contribute to network diversity [@bargsted_pautas_2020; @plaza_juntos_2022;@lazarsfeld_friendship_1954]. In this sense, socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004;@homans_human_1951]. Nonetheless, it can be argued that attitudinal similarity is the result of segregated social relations, which in turn are linked to structural processes of differentiation that also demarcate the boundaries to form ties with other social positions [@feld_focused_1981].

It is possible to find two approaches in the study of class relations from a network approach. On the one hand, it has emphasized the role of _diversity_ as the degree of connectedness to dissimilar occupations that vertically represent access to resources embedded in social networks [@lin_social_2007]. However, diversity is defined as the rate of dissimilar ties within the network that do not necessarily count with a reference position to describe the network composition. A second approach has underscored the role of _segregation_ as the lack of cross-class network ties. This perspective is conceptually closer to homophily as it is anchored in individual class positions and has empirically been addressed through the concept of network homogeneity [@otero_open_2021].

Empirically, studies on network diversity have shown that higher civic engagement in formal organizations increases the chances of bridging with diverse people among the upper class, in contrast to the more homogeneous participation of the working class [@pichler_social_2009]. Similar patterns have been found in terms of the composition of social ties, where the upper and intermediate classes hold increasingly diverse and prestigious social environments than the working classes [@cepic_how_2020;@carrascosa_class_2023]. Also, this stratified network structure holds during the life course, where the upper classes hold increasingly diverse social contacts in contrast to the stable networks of the working classes [@volker_social_2020]. In contrast, studies that have focused on segregation suggest that the property dimension is much less permeable than authority-based boundaries, arguing that class interests increase the social distance between proprietors and manual workers, while the intermediate class position of supervisors jointly to their higher contact frequency with manual workers make friendship tie formation more likely [@wright_relative_1992]. Similar studies have shown that the intermediate class holds higher permeability in contrast to the more homogeneous networks of the working class, suggesting that their limited life chances and lower capacity to be socially engaged ultimately result in a lack of social resources that lead to social segregation. In contrast, the upper class is less permeable and homogeneous because it tends to self-select as a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

Besides individual class-based mechanisms, it can be argued that their social relations represent a relevant factor in attitude formation. Particularly, despite limited research on the link between social networks and redistributive preferences, two theoretical approaches have discussed the role of social relations in attitude formation: reference groups and class-based networks.

It has been stated that perceptions about economic inequality rooted in social comparison processes with reference groups can explain the formation of redistributive preferences [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. The argument posits that people form their beliefs through family, friends, and coworkers' experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of these reference groups [@Evans1992, p. 467]. Therefore, inferences about the social world are linked to the degree of segregation in the immediate social environment of a person, which influences the intensity and character of the information that ultimately shapes inequality perceptions [@mijs_is_2021]. Accordingly, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more segregated networks [@summers_deliberating_2022]. Nevertheless, I argue that this body of research has been mainly focused on the cognitive dimension of preference formation through inequality perceptions or beliefs rather than straightforward addressing the claimed influence of network segregation on redistributive preferences [@cansunar_who_2021;@garcia-castro_perceived_2022;@Cruces2011;@becker_significant_2021].

In contrast, a network approach provides a better picture of class relations that nurture social norms and group identity [@kalmijn_social_2007, p. 550]. More precisely, it has been argued that redistributive preferences are influenced by the class situation of both the individual and network ties, where social influence processes can either allineate or divide opinions according to the contacts' class position and how segregated contact opportunities are [@vargassalfate_contact_2023]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to individual and household material well-being, cultural perspectives, and political preferences that structure broader social experiences [@morris_attenuation_1996, p. 48]. Furthermore, social integration can be affected in societies with lower contact opportunities between different social classes, creating an "empathy gulf" that comprises barriers to imagining others' lifestyles in contexts of rising inequality [@sachweh_moral_2012]. Thus, spatially segregated interactions may nurture doubts about worthiness when the lower classes contrast themselves with the lifestyles of the upper classes, which in turn can undermine feelings of social inclusion and cohesion [@sachweh_moral_2012]. As a result, segregation drives the lives of others to become more distant and might have consequences for empathy and solidarity toward others, potentially leading to the perception of fellow citizens as strangers [@otero_lives_2022, p. 758].

<!-- The attention has increased in recent years to how social ties are connected to attitudes toward inequality, and it has been shown that frequent interactions between classes can lead to different attitudinal changes depending on the individuals' social class. For instance, contact with the upper classes drives higher justification of economic inequality among the lower classes, whereas interactions with lower classes may prompt the different social classes to question the fairness of income distribution [@vargassalfate_contact_2023]. Further evidence suggests that class-based contact diversity shortens the social distance between social classes, providing broader images of the living conditions of others, which can trigger greater concerns about inequality [@otero_power_2023] and undermine support for the market distribution of social services [@beck_explorando_2019]. -->

The class position of surrounding family members, friends, and acquaintances influences redistributive preferences as they function as socialization agents that can be bolstered in segregated social networks. In principle, political attitudes are connected to class interests and norms as they are nurtured in the family of origin during childhood and early adulthood. For instance, it has been shown that individuals with network ties to the upper class through parental connections support redistribution and progressive taxation less than those from working-class family backgrounds [@lee_consider_2023]. Also, as households share risk according to the class position of their members, redistributive preferences are not only affected by their family background but also by partners' class positions. For example, @paskov_crossclass_2022 found that working-class ties bolster redistributive preferences but decrease with upper-class ties, becoming much more intense when the class positions of individuals, partners, and parents form a more homogeneous network. Besides, @lindh_missing_2021 found that including friends and acquaintances in the network structure leads individuals with higher ties to the managerial class to decrease their redistributive preferences compared to those with higher sociocultural and working-class ties, suggesting that people tend to consider and adjust their attitudes according to the class position of their contacts [@lindh_missing_2021, p. 698].

In summary, the tendency to form homogenous social ties regarding social class is expected to reinforce attitudes and consolidate opinion similarity. Empirically, a weak direct association between homogeneity and redistributive preferences is expected because class segregation does not distinguish between classes as it provides only the _overall_ degree of segregation. In contrast, the focus is on how class-based network homogeneity is conditional to social class. Empirically, this can be observed through the interaction between network homogeneity and social class. Precisely, this will indicate if being segregated into lower (upper) classes leads to stronger (weaker) redistributive preferences ($H_1$).

## Economic inequality as context for class relations and redistributive preferences

Studies that have stressed the role of economic inequality on the relationship between social class and redistributive preferences indicate that the lower and the upper classes react differently to rising economic inequality. Theoretically, political economists have suggested that high-income individuals are far from monolithic in their redistributive preferences, arguing that their concerns about the harmful _consequences_ of economic inequality (e.g., crime) ultimately motivate altruistic support for redistribution [@rueda_food_2018;@dimick_models_2018;@dimick_altruistic_2017;@rueda_externalities_2016]. Conversely, the moral economy literature in sociology has argued that the differences among the affluent can be explained as a matter of distributive justice evaluations about the _procedures_ for resource allocation [@liebig_sociology_2016;@atria_economic_2020]. Hence, affluent groups are more responsive because they perceive that increasing inequality harms the overall opportunity structure and social mobility chances [@sachweh_criticizing_2017]. Likewise, lower perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of justice in the conditions for getting ahead [@kim_socioeconomic_2018]. In contrast, low-income individuals perceive ascribed characteristics as more important in constraining the opportunity structure, regardless of current income inequality [@sachweh_why_2019, p. 656].
